Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2008-12-01 docs/pubs/ISEI6/main.tex

\begin{abstract}
  Formal ontlogies promise to provide many added advantages to
  ecological data management ... Most ontology editing frameworks and
  tools, however, are design for knowledge-representation experts and
  those with considerable understanding of the complexities of
  ontology languages (i.e., description logics). This paper proposes
  an approach for ontology creation based on ...
\end{abstract}

@@

The use of formal ontologies can significantly enhance metadata
descriptions of ecological data by providing richer and more
consistent terminologies \cite{tree,arees,parr}. Annotating data with
ontology terms can both help users interpret data as well as enable
advanced capabilities for data discovery and integration, e.g., by
exploiting subsumption and part-of hierarchies as well as more formal
constraints such as cardinality restrictions on properties and term
equivalence \cite{ecoinf,other?}. Many efforts are underway to engage
scientists in the development of ontologies to represent common
ecological and biodiversity concepts. Most of these projects leverage
the W3C Web Ontology Language (OWL) as a standard XML syntax for
representing and sharing ontologies. A key advantage of OWL is that it
is supported by a wide range of generic tools, including editors
\cite{protege,swoop}, reasoning systems \cite{pellet,fact++}, query
languages \cite{???}, and storage technologies \cite{jena,???}. Most
of these tools, however, are primarily targeted at experts in
knowledge engineering and software development. This is especially
true with ontology editors (such as Protege, OWLed, SWOOP, etc.),
which allow for very detailed ontology specifications, but at the same
time require a considerable amount of understanding of the underlying
ontology formalisms and syntax. Thus, we see the lack of suitable
ontology editing tools for scientists as one of the major barriers for
more wide-scale adoption of ontologies in ecology.

>>>>

\begin{abstract}
  Data discovery and integration is important in many ecological
  studies, especially those that concern broad-scale ecological
  questions. While important, the discovery and integration of
  ecological data is often a complex and time-consuming task for
  researchers. This situation is often due to the use of informal,
  ambiguous, and sometimes inconsistent terms for describing
  ecological data.  Ontologies can help address this problem by
  providing a mechanism to formally define ecological concepts that
  can be used to more consistently annotate, relate, and search for
  data sets.

  Unlike in molecular biology or biomedicine, however, few ontology
  development efforts exist within ecology. These efforts often
  require considerable expertise in ontology languages and development
  tools that limits the creation of ontologies by ecologists.  This
  paper addresses this problem by allowing ecologists to use common
  spreadsheet tools to describe different aspects of an ontology. We
  introduce novel conventions for creating, relating, and constraining
  concepts through spreadsheets, and provide software tools for
  converting ontologies created in this way into equivalent OWL-DL
  representations. We also consider inverse translations, i.e., that
  convert ontologies represented using OWL-DL into our spreadsheet
  format. Our approach allows large lists of terms to be easily
  related and organized into concept hierarchies, and generally
  provides a more intuitive and natural interface for ontology
  creation by ecologists.
\end{abstract}

@@

Within the fields of molecular biology and biomedicine considerable
effort has gone into developing ontologies for improving data
discovery and integration
\cite{bard04:_ontol_in_biolog,ashburner00:_gene_ontol}. While similar
benefits can be obtained for ecological data, far fewer efforts exist
to develop richer and more consistent terminologies of ecology
concepts
\cite{madin08:_advan_ecolog_resear_with_ontol,parr20:_data_sharin_in_ecolog_and_evolut}.
The use of formal ontologies can significantly enhance metadata
descriptions of ecological data. Annotating data with ontology terms
can both help users interpret data as well as enable advanced
capabilities for data discovery and integration, e.g., by exploiting
subsumption and part-of hierarchies as well as more formal constraints
such as cardinality restrictions on properties and term
equivalence. Many efforts are underway to engage scientists in the
development of ontologies to represent common ecological and
biodiversity concepts. Most of these projects leverage the W3C Web
Ontology Language (OWL) \cite{smith04:_owl_web_ontol_languag_guide} as
a standard XML syntax for representing and sharing ontologies. A key
advantage of OWL is that it is supported by a wide range of generic
tools, including editors
\cite{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\cite{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\cite{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\cite{carroll04:_jena,broekstra02:_sesam}. Most of these tools,
however, are primarily targeted at experts in knowledge engineering
and software development. This is especially true with ontology
editors (such as Protege, SWOOP, etc.), which allow for very detailed
ontology specifications, but at the same time require a considerable
amount of understanding of the underlying ontology formalisms and
syntax. Thus, we see the lack of suitable ontology editing tools for
scientists as one of the major barriers for more wide-scale adoption
of ontologies in ecology.
